Wine Quality
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(car)
library(reshape2)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityReds.csv')
```
```{r echo=FALSE, Looking_at_the_data}
str(wine)
summary(wine)
```

The wine dataset contains 13 variables and 1599 observations. I will look at the
factors that affect the wine quality.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(4,16,1))
```

```{r echo=FALSE}
summary(wine$fixed.acidity)
```

Most of the wine has fixed acidity quantity around 7 g/dm^3. It would be
interesting to see how the quantity of fixed acidity fair with the quality of
the wine.

```{r echo=FALSE}
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.02)
```

It's a bi-modal distribution. We will have to see if this type of distribution
has an impact on the quality.

```{r echo=FALSE}
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(binwidth = 0.001) +
  scale_y_continuous(breaks = seq(0, 150, 25))
```

```{r echo=FALSE}
range(wine$citric.acid)
```

Now, this is interesting. As citric acid is added in small quantity as it can
add freshness and flavour to the wine. At least one-fourth of the wine doesn't
have citric acid. I would be interesting to see the quality of wine for those
with citric acid as Zero.

```{r echo=FALSE}
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(binwidth = 0.1)
```

A normal distribution with lots of Outliers. It would be interesting to know
about the outliers quality.

```{r echo=FALSE}
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(binwidth = 0.001)
```

Sodium chloride is a salt. Most of the wine has less than 0.1 g/dm^3. With few
outliers. I wonder what would be the quality of the wine which has chloride
quantity more than 0.6 g/dm^3.

```{r echo=FALSE}
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 0.1)
```

Free sulfur dioxide data is skewed. Should we transform the data?

```{r echo=FALSE}
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = .5)
```

Both free sulfur dioxide and total sulfur dioxide's data is skewed. How strong is
Is it correlation with the free sulfur dioxide?

```{r echo=FALSE}
ggplot(aes(x = density), data = wine) +
  geom_freqpoly(color = 'blue', binwidth = 0.001)
```

Does density have any effect on the quality score of the wine?

```{r echo=FALSE}
ggplot(aes(x = pH), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = seq(2.5,4.2,.1))
```

A bulk of the wine has a pH between 3.15 and 3.45. I wonder what this plot might
look like across the quality of the wine. Does the pH have any effect on the
quality of the wine? There are a few outliers, it would be interesting to know
about their quality too.

```{r echo=FALSE}
ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(binwidth = 0.01)
```

Sulphates are used as antioxidant, I think it might have a strong correlation with
the quality.

```{r echo=FALSE}
ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(binwidth = .01)
```

Little skewed. It would be nice to see how it fairs with the quality of the
wine. A lot of wine have below 10. How does the below 10 and above 10 vary with
quality?

```{r echo=FALSE}
ggplot(aes(x = quality), data = wine) +
  geom_bar()
```

The data in quality is an integer. A bulk of the rating is either 5 or 6. Will
be interesting to know whether the variables have the same observation where the
quality is 5 or 6.

```{r}
# Create a categorical variable
wine$quality_category <- cut(wine$quality, c(3,5,8))
```

I created a categorical variable that will help in visualizing in a more
appropriate way.

```{r echo=FALSE}
ggplot(aes(x = quality_category), data=subset(wine, !is.na(quality_category))) +
  geom_bar()
```

So, most of the wines fit into the 5-8 category.

# Univariate Analysis

### What is the structure of your dataset?

The dataset has 13 variables and 1599 observations. Variables are the chemical
property of the wine. Most of the variables are normally distributed. There was
one variable with bi-modal distribution - Volatile Acidity. Free sulfur dioxide
and total sulfur dioxide data is skewed. And most of the variables do have
outliers too.

### What is/are the main feature(s) of interest in your dataset?

How the Alcohol effects the quality would be interesting to know. Which chemical
is strongly correlated with the quality? and as we saw that one-fourth of the
wine do not have citric acid, I will have to see their quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

pH level in the best quality of alcohol will be worth observing. Outliers in
most of the variables would be worth investigating.

### Did you create any new variables from existing variables in the dataset?

Yes, I created a quality_category where it categories the quality score. I
divided the score into two categories - 3-5 and 5-8.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Some plot didn't look like a normal distribution with default binwidth, so I
adjusted the binwidth which gave a better shape to the plot
(normal distribution). Adjusting the binwidth also helped me in getting a
bi-modal distribution which looked like a normal distribution with default
binwidth.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, Bivariate_Plots, fig.width=12, fig.height=12}
# Plot matrix
ggpairs(wine,
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'))),
        columns = c('fixed.acidity','volatile.acidity',
                                'citric.acid','free.sulfur.dioxide',
                                'total.sulfur.dioxide','density','pH',
                                'sulphates','alcohol','quality'))
```

As per the plot matrix, we can see the correlation between variables. Some of
the interesting correlation is between alcohol and quality, alcohol and density
and volatile acidity.

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_jitter(alpha = 1/15) +
  scale_x_continuous(breaks = seq(8,15,.5))
```

I added jitter for better visual.

From the above scatterplot, we can say that alcohol percentage greater than 9.5
might result in as a good quality wine. However, the percentage of alcohol
solely doesn't result in a better quality.

```{r echo=FALSE}
ggplot(aes(x = density, y = alcohol), data = wine) +
  geom_point()
```

Density and alcohol correlation was a stronger one in our plot matrix and we can
see it here.

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = pH), data = wine) +
  geom_point()
```

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = pH), data = wine) +
  geom_point(color = 'orange') +
  geom_smooth(method = 'lm') +
  coord_cartesian(xlim = c(5.5, 13))
```

Re-plotting the above plot. Like our earlier finding in plot matrix they do have
a strong correlation.

```{r echo=FALSE}
ggplot(aes(x = quality, y = citric.acid), data = wine) +
  geom_jitter(alpha = 1/15)
```

Not a single wine got a quality score as 8, when the citric acid is 0.

```{r echo=FALSE}
ggplot(aes(x = volatile.acidity, y = quality), data = wine) +
  geom_jitter(alpha = 1/15)
```

Lesser the volatile acidity better the quality.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Citric acid and quality of the wine are not strongly correlated. At first, I
thought it might be a big factor. Quality is correlated with the alcohol, pH has
a correlation with the citric acid and fixed acidity. The quality is correlated
with the percentage of alcohol. It would be interesting to see what percentage
of alcohol makes a wine of good quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Residual sugar doesn't have a strong correlation with any variable. So I removed
it from the plot matrix.

### What was the strongest relationship you found?

The strongest relationship was between fixed acidity and pH with -0.683.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_plots}
ggplot(aes(x = alcohol, y = volatile.acidity),
       data = subset(wine, !is.na(quality_category))) +
  geom_point(aes(color = quality_category))
```

This plot strengthens out previous assumption - if the alcohol percentage it
more then, the quality might be better.

```{r echo=FALSE}
ggplot(aes(x = alcohol, y = density),
       data = subset(wine, !is.na(quality_category)))+
  geom_point(aes(color = quality_category))
```

It justifies our earlier finding in the plot matrix, where there was a strong correlation between the density and the alcohol and although there is a strong correlation between alcohol and the quality it doesn't necessarily mean the
density and quality have a strong correlation.

```{r echo=FALSE}
ggplot(aes(x = citric.acid, y = fixed.acidity),
       data = subset(wine, !is.na(quality_category)))+
  geom_point(aes(color = quality_category))
```

The quality category seems to be all over the place. There doesn't seem to be
any strong correlation between the quality and the other two variables. However
the fixed acidity and citric acid are strongly correlated.

```{r echo=FALSE, linear_model}
m1 <- lm(I(quality) ~ I(alcohol), data = wine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + citric.acid)
mtable(m1, m2, m3, m4, sdigits = 3)
```

By looking at the R^2 (coefficient of determination) 33.6% it indicates that the
model (m1, m2, m3, m4) does explain the variability in the quality of the wine.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The two clear variables that do contributes to the quality of wine are the
percentage of alcohol and volatile acidity. Other variables don't strongly
explain the variability in the quality of the wine.

### Were there any interesting or surprising interactions between features?

Volatile acidity had a strong correlation with citric acid, and as volatile
acidity explains the variability of the quality, I thought citric acid will also
have an impact on the quality but that wasn't to be.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I did create models, it helped me justify the variability of quality of
the wine due to various variables.

Strengths:

Very simple to use and comprehend the result.

Limitations:

Linear regression is very sensitive to the anomalies in the data.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# To centre the title
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(aes(x = volatile.acidity, y = quality), data = wine) +
  geom_jitter(alpha = 1/15, color = 'red') +
  scale_x_continuous(breaks = seq(0.12, 1.58, .1)) +
  xlab("Volatile Acidity") +
  ylab("Quality") +
  ggtitle("Volatile Acidity vs Quality")
```

### Description One

We can clearly see that wine with volatile acidity between 0.22 and 0.92 have a
good quality score. 

### Plot Two

```{r echo=FALSE, Plot_Two}
ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_jitter(colour="blue", alpha = 1/15) +
  scale_x_continuous(breaks = seq(8,15,.5)) +
  xlab("Alcohol") +
  ylab("Quality") +
  ggtitle("Alcohol vs Quality")
```

### Description Two

This plot (quality vs alcohol) gives an insight about the quality of the wine -
Greater the percentage of alcohol greater the quality.

### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = volatile.acidity),
       data = subset(wine, !is.na(quality_category))) +
  geom_point(aes(color = quality_category)) +
  scale_x_continuous(breaks = seq(8.5,15,1)) +
  xlab("Alcohol") +
  ylab("Volatile Acidity") +
  ggtitle("Volatile Acidity and Alcohol with Quality Category")
```

### Description Three

Alcohol and Volatile acidity are the two variables which explain the 31.7% of
the variability in the quality score. In our previous findings, we saw how the
volatile acidity had a negative correlation which can be seen in this plot,
lesser the volatile acidity better the quality. Similarly, with the alcohol, the
only difference is that it is positively correlated with the quality.

Note: We can only infer the above statement. Having a low volatile acidity and a
high percentage of alcohol doesn't necessarily means the quality will be good.
There are other factors that can have an effect on the quality of the wine.

------

# Reflection

I started exploring by taking a look at the data and there types most of them
were numbers except one which was an integer. Then I had a look at the
descriptive statistics which gave an idea of the spread, mean, median and the
quartile range. Understanding the variables help me in exploring the variables
with the plot. Initially, I went with the uni-variable plots along the course I
put up some questions which helped me in exploring in the later stage. Most of
the variables were have a normal distribution and a few were skewed. I did
encounter outlier which was surprising as well. I did create a categorical
variable which grouped the quality into two groups, the quality of wine from 3
to 5 were in one group and 5 to 8 were in a different group. This variable
helped me in getting deeper insights into the later stage of the exploration.

With the plot matrix, I got a clear idea of how the variables are related to one another. This matrix showed how alcohol was strongly correlated to the quality
of the wine, it also helped me in selecting the variables of my interest which
saved a lot of time as exploring every variable could have been tedious. I
plotted some scatter plot between two variables to visualize how the variables
are correlated.

In the multivariate analysis, it visualized with the categorical variable which
I created in the earlier part of my exploration. The categorical variable helped
in getting an insight of the quality of wine while exploring different variables
at once. Finally, I created a linear model to predict the quality of the wine
across different variables. Alcohol and Volatile acidity account for 31.7% of
the variation in the quality. One of the most surprising things was when I found
out that citric acid didn't improve the quality of the wine.

As all the variables were either numbers or integers I initially struggle to
create a categorical variable so that I could explore further - I was finding it
hard to think of a categorical variable. Eventually, I did it using the quality variable.

Further, I could have explored the outlier as there were outliers in most of the variables.

Limitations in our exploration: As with our liner model I could only justify the
31.7% of variation in the quality by two variables if I would have had some more
details about of the wine I could have done more in finding out the variability
of the quality, The data was described in 2009 and as things changes over the
time today this correlation of alcohol and volatile acidity might not be the
same in today's world.

# Reference

* ggpair documentation for plot matrix [here](https://www.rdocumentation.org/packages/GGally/versions/1.4.0/topics/ggpairs).

* About [Linear Model](https://www.google.co.in/url?sa=t&rct=j&q=&esrc=s&source=web&cd=2&cad=rja&uact=8&ved=0ahUKEwiw6IeRt7fcAhXLsY8KHQ0GDRwQFggpMAE&url=https%3A%2F%2Fen.wikipedia.org%2Fwiki%2FLinear_model&usg=AOvVaw38KON0TNfgRAWFJiKDQQAq).

* About [Correlation](https://www.surveysystem.com/correlation.htm).

* ggplot documentation [here](https://www.rdocumentation.org/packages/ggplot2/versions/3.0.0).